---
title: "META monthly data analysis"
author: "Lynnstacy Kegeshi"
date: "2022-12-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE}
library(dplyr)
library(lubridate)
library(xts)
library(tseries)
library(forecast)
library(seastests)
library(FinTS)
library(tidyverse)
library(ggplot2)
```

# Preliminary Analysis
We shall be using the META monthly stock price data in our time series analysis.The read.csv function reads the data in as a data frame, and assign the data frame to a variable. Then we check the structure of the data to see the data types in our data frame
```{r}
metam<-read.csv("META monthly historical prices data.csv", header = TRUE,  sep = ",")
str(metam)
```

The date variable is a character. We need to change it to a date format in before plotting our time series

```{r}
metam$New_date=ymd(metam$Date)
str(metam$New_date)
```
This command creates a new column New_Date and the data type in this column is in date format.

## Plotting the time series
We chose to use the Closing price column setting the frequency to 12, monthly. 

```{r}
colnames(metam)
closingts<-ts(metam$Close,start=c(2012,6),end=c(2022, 5), frequency = 12)
autoplot(closingts,main = "Closing price TS", xlab = "Time in years", ylab="Closing price" )


```

The graph above is a time series of the closing price from 2012 to 2022

### Box plot of the closing price
In order to compare the distribution of the closing price between the months we used a boxplot
We first needed to extract the months from the data
```{r}
metam$monthhh=month(ymd(metam$New_date), label = TRUE)
metam$monthhh
```

After that we plot it
```{r}
boxplot(Close~monthhh, 
        data=metam, 
        main="Box plot showing META closing price for each month",
        xlab="Months",
        ylab="Closing price in US Dollars",
        col="deepskyblue",
        border="darkblue")
```

##Decomposing the time series
```{r}
closing_dec<-decompose(closingts, "multiplicative")
closing_dec
plot(closing_dec)
```
The time series has an upward trend.
The variability in the data is also increasing over time.
### Plot the original ACF AND PACF 
```{r}
#Plot the original data ACF and PACF
ggtsdisplay(closingts)
```

The ACF plot shows slow decay of lag to 0 indicating an AR model. The PACF plot suggests AR model of the order 1 AR(1) as PACF number is close to 0 after lag 1.

### Test for stationarity
#### ADF test
```{r}
adf.test(closingts)
```

pvalue>0.05: 0.07992>0.05, Fail to reject the null hypothesis and conclude that the time series is non stationary

#### KPSS test
```{r}
kpss.test(closingts)
```
p-value is 0.01. 0.01<0.05, Reject null hypothesis and conclude that the time series is non stationary.

## Make the data stationary
We need to remove the upward trend of the data and the variability in order to stationarize this time series.

### Removing Variability Using Logarithmic Transformation
Since the data shows changing variance,log transformation of the data using the log() function will stabilize the data. The resulting series will be a linear time series.
check how many times we need to differentiate the data in order to make it stationary

```{r}
closingts_linear<-log(closingts)
```

Plotting the resulting series
```{r}
plot.ts(closingts_linear, main="Closing Stock Prices(log)", ylab="Closing Price", col=4)

```
Log transformation of the original time series has enabled more “balanced” variance observed throughout the time series, and in a way coerced the data into stationarity.

### Removing Linear Trend
Check how many times we need to differentiate the log transformed time series

```{r}
ndiffs(closingts_linear)
```
We will now perform the first difference of the log transformed series to make it stationary

Differentiate it and plot the time series, ACF AND PACF plot
```{r}
diffclose<-diff(closingts_linear)
diffclose
plot.ts(diffclose, main="Change in closing price per month",xlab="Time in years", ylab="Closing price", col="blue")
ggtsdisplay(diffclose)


```

based on our limited analysis, let's say we will go with p=0, d=0, and q=0.

Our suggested model is then ARIMA(0,0,0).

Now check number of differences in the differenced time series for confirmation that the time series has been detrended.
```{r}
ndiffs(diffclose)
```

0 indicates we can't differentiate the time series any further.

### Seasonality and trend
```{r}
isSeasonal(diffclose, test="wo")
```

The time series shows no evidence of seasonality.




##### ADF test
```{r}
adf.test(diffclose)
```
p-value is 0.01. 0.01<0.05, we thus reject the null hypothesis that the time series is non stationary and conclude that the time series is stationary.

##### KPSS test
```{r}
kpss.test(diffclose)
```

0.1>0.05 fail to reject null hypothesis and conclude that the time series is stationary




## ARIMA 
ARIMA models cannot handle any type of non-stationarity and the time series is now stationary so we can proceed.
To find the best ARIMA model to fit our time series, we;
```{r}
ARIMAfitmeta <- auto.arima(diffclose, trace=TRUE,
                       ic= c("bic"), #c("aicc","aic","bic"),
)
ARIMAfitmeta
```
The best model is thus ARIMA(0,0,0)

Next we fit the model as ARIMA(0, 0, 0).
```{r}
diffclose_ar<- arima(diffclose, order=c(0,0,0))
diffclose_ar
BIC(diffclose_ar)

```

