---
title: "META monthly data analysis"
author: "Lynnstacy Kegeshi"
date: "2022-12-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE}
library(dplyr)
library(lubridate)
library(xts)
library(tseries)
library(forecast)
library(seastests)
library(FinTS)
library(tidyverse)
library(ggplot2)
```

# Preliminary Analysis
We shall be using the META monthly stock price data in our time series analysis.The read.csv function reads the data in as a data frame, and assign the data frame to a variable. Then we check the structure of the data to see the data types in our data frame
```{r}
metam<-read.csv("META monthly historical prices data.csv", header = TRUE,  sep = ",")
str(metam)
```

The date variable is a character. We need to change it to a date format in before plotting our time series

```{r}
metam$New_date=ymd(metam$Date)
str(metam$New_date)
```
This command creates a new column New_Date and the data type in this column is in date format.

## Plotting the time series
We chose to use the Closing price column setting the frequency to 12, monthly. 

```{r}
colnames(metam)
closingts<-ts(metam$Close,start=c(2012,6),end=c(2022, 5), frequency = 12)
autoplot(closingts,main = "Closing price TS", xlab = "Time in years", ylab="Closing price" )


```

The graph above is a time series of the closing price from 2012 to 2022

### Box plot of the closing price
In order to compare the distribution of the closing price between the months we used a boxplot
We first needed to extract the months from the data
```{r}
metam$monthhh=month(ymd(metam$New_date), label = TRUE)
metam$monthhh
```

After that we plot it
```{r}
boxplot(Close~monthhh, 
        data=metam, 
        main="Box plot showing META closing price for each month",
        xlab="Months",
        ylab="Closing price in US Dollars",
        col="deepskyblue",
        border="darkblue")
```

##Decomposing the time series
```{r}
closing_dec<-decompose(closingts, "multiplicative")
closing_dec
plot(closing_dec)
```
The time series has an upward trend.

### Plot the original ACF AND PACF 
```{r}
#Plot the original data ACF and PACF
ggtsdisplay(closingts)
```

### Test for stationarity
#### ADF test
```{r}
adf.test(closingts)
```

pvalue>0.05: 0.07992>0.05, Fail to reject the null hypothesis and conclude that the time series is non stationary

#### KPSS test
```{r}
kpss.test(closingts)
```
p-value is 0.01. 0.01<0.05, Reject null hypothesis and conclude that the time series is non stationary.

### Make the data stationary
check how many times we need to differentiate the data in order to make it stationary

```{r}
ndiffs(closingts)
```
We need to differentiate the data only once to make it stationary

Differentiate it and plot the ACF AND PACF 
```{r}
dfclosing<-diff(closingts)
dfclosing
autoplot(dfclosing, main="Change in closing price per month",xlab="Time in years", ylab="Closing price", col="blue")
ggtsdisplay(dfclosing)


```

Now check if the time series is stationary after differencing.
```{r}
ndiffs(dfclosing)
```

0 indicates we can't differentiate the time series any further.

##### ADF test
```{r}
adf.test(dfclosing)
```
p-value is 0.1257. 0.1257>0.05, we thus fail to reject null hypothesis and conclude that the time series is non stationary.

##### KPSS test
```{r}
kpss.test(dfclosing)
```

0.1>0.05 fail to reject null hypothesis and conclude that the time series is stationary

Since the series is stationary according to the KPSS test but is not stationary according to the ADF test, we conclude that the series is stationary around a deterministic trend and so is fairly easy to model this series and produce fairly accurate forecasts.

### Seasonality and trend
```{r}
isSeasonal(closingts, test="wo")
```

The time series shows no evidence of seasonality.



